Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added first draft of summary
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2911 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8ffe9f07bc007fff7273a37fdb0da500017ae494

%To evaluate the proposed circular regression trees and forests, a case study on
%probabilistic forecasting of wind direction for two airports in Austria is
%presented. In addition to the tree-based models, alternative common approaches
%are used as a benchmark in order to assess the predictive performances of
%circular regression trees and forests. 
}
\Keywords{Distributional Trees, Random Forests; Circular Data; von Mises Distribution}

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Circular regression trees and forests for circular responses are established by coupling 
model-based recursive partitioning with the von Mises distribution.

The resulting 
trees can capture non-linear changes, shifts, and potential interactions in covariates 
without prespecification of such effects. This is particularly useful for modeling a 
highly fluctuating response such as wind direction. \dots

\begin{itemize}
  \item Add summary/conclusion of case study;
  \item Add summary/conclusion of circular regression trees and forests.
  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
\end{itemize}

>>>>

%To evaluate the proposed circular regression trees and forests, a case study on
%probabilistic forecasting of wind direction for two airports in Austria is
%presented. In addition to the tree-based models, alternative common approaches
%are used as a benchmark in order to assess the predictive performances of
%circular regression trees and forests. 
}
\Keywords{Distributional Trees; Random Forests; Circular Data; von Mises Distribution}

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and forests 
are established by coupling model-based recursive partitioning with the von Mises distribution. 
By separating the data into more homogeneous subgroups, possible difficulties in circular regression
are bypassed as covariates are solely considered for splitting and 
group-specific models are fitted as intercept-only models.
%distributional models are fitted as intercept-only models in each node. 
In addition, by specifying the von Mises distribution for each node and
allowing for splits in both distribution parameters $\mu$ and $\kappa$, 
fully probabilistic forecasts are provided.

The performance of the novel circular trees and forests are assessed in a real case study for shortterm
probabilistic wind direction forecasting at two airports with different site characteristics.
As benchline models, probabilistic climatology and perstistency models, as well as a state-of-the-art
circular GLM are evaluated based on proper scoring rules.
In summary, the circular trees and forests have the highest predictive performance in this setting.
For cases without changes in the wind regime, lagged response values provide already
highly skillful estimates leading to good performance of the persistency model for shortterm wind direction forecasts. While in these cases the trees and forests also benefit from the highly informative lagged response, they can compensate for a lower information of this covariate by incorporating other quantities and possible interactions of these, contrary to the persistency model (see Figure~\ref{fig:boxplot_crpsskill}).
Hence, the tree-based models provide reliable forecasts in all tested meterological settings.
For operational use, a possible extension could be the incorporation of numerical weather predictions as (additional) covariates. While this probably only slightly improves the predictive skill for short leadtimes, it possibly extends the potential forecast range of the different methods.

% TODO: circular splits (see first version of IWSM, 4 pages)

To conlcude, the tree structure can capture non-linear changes, shifts, and potential interactions in 
covariates without prespecification of such effects. As supported by the presented case study,
this can be particularly useful for modeling a highly fluctuating response, such as typically 
observed for wind direction, or/and in case of a large number of possible covariates.
Moreover, the case study shows that building ensembles of circular trees can even improve the forecasting performance, as the resulting circular forests allow for modeling smooth effects and stabelize the model.

%The comparison of the novel circular trees with simple climatological and persistency models, depicted
%for probabilistic wind direction forecasts at two airports, shows that even in case of no obvious 
%skillful covariates the trees have an high predictive performance.

%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{itemize}
%  \item Add summary/conclusion of case study;
%  \item Add summary/conclusion of circular regression trees and forests.
%  \item Outlook on possible extensions (use of NWP data, circular splits, ...).
%\end{itemize}
